### 2023-04-22 chapter2/chapter2.tex
Re-worded motivation for use of contrasts

To better understand the relationship of this variability to differences in the regions' biophysical properties, we sought to correlate the average disorder score of the segments in a region with their molecular features. However, because the sequences are evolutionarily related, they are not independent, making them unsuitable for direct use in many standard statistical procedures. We instead applied the method of contrasts, which takes differences between adjacent nodes in the phylogenetic tree relating the species to generate a set ``contrasts,'' to both the scores and the features. Under some general assumptions of the underlying evolutionary process, contrasts are independent and identically-distributed and are therefore appropriate for use in correlation analyses. The resulting feature contrasts have varying degrees of correlation with the score contrasts (Fig.~\ref{fig:scores}B-C). Some, like isopoint, are uncorrelated, but most are significantly, if weakly, correlated. In general, the strongest correlations are observed for features which have a direct biophysical relationship to the presence or absence of disorder, such as disorder\_fraction or hydrophobicity. Interestingly, the correlations with many motifs were statistically significant, though small in magnitude relative to the non-motif features. However, a more detailed analysis of this observation is presented in the discussion. To determine if regions with rapidly evolving disorder scores were associated with a particular function, process, or compartment, we then extracted the upper decile of their rate distribution and performed term enrichment analyses on their associated annotations (Fig.~\ref{sfig:scores_histogram}). The most significant terms are generally related to DNA repair or extracellular structure, which suggests these processes and components are enriched in proteins whose structural state is rapidly evolving (Fig.~\ref{fig:scores}E).

>>>>

To better understand the relationship of this variability to differences in the regions' biophysical properties, we sought to correlate the average disorder score of the segments in a region with their molecular features. However, as the sequences are not independent but instead related by a hierarchical structure which reflects their evolutionary relationships, any derived features are unsuitable for direct use in many standard statistical procedures. In the most severe cases, traits derived from clades of closely-related species can effectively act as duplicate observations, which can yield spurious correlations. We therefore applied the method of contrasts, which takes differences between adjacent nodes in the phylogenetic tree relating the species to generate a set of ``contrasts,'' to both the scores and the features. Under some general assumptions of the underlying evolutionary process, contrasts are independent and identically-distributed and are therefore appropriate for use in correlation analyses. The resulting feature contrasts have varying degrees of correlation with the score contrasts (Fig.~\ref{fig:scores}B-C). Some, like isopoint, are uncorrelated, but most are significantly, if weakly, correlated. In general, the strongest correlations are observed for features which have a direct biophysical relationship to the presence or absence of disorder, such as disorder\_fraction or hydrophobicity. Interestingly, the correlations with many motifs were statistically significant, though small in magnitude relative to the non-motif features. However, a more detailed analysis of this observation is presented in the discussion. To determine if regions with rapidly evolving disorder scores were associated with a particular function, process, or compartment, we then extracted the upper decile of their rate distribution and performed term enrichment analyses on their associated annotations (Fig.~\ref{sfig:scores_histogram}). The most significant terms are generally related to DNA repair or extracellular structure, which suggests these processes and components are enriched in proteins whose structural state is rapidly evolving (Fig.~\ref{fig:scores}E).
